Customizing Software by Direct Manipulation of Structured Data

### 2020-05-20 paper.tex
better architecture intro

  We describe Wildcard, a browser extension which demonstrates
  table-driven customization in the context of web applications. Through
  concrete examples, we demonstrate that this paradigm can be used to
  create useful customizations for a variety of real applications. We
  share reflections from our usage of the Wildcard system, including its
  strengths and limitations relative to other customization approaches.
  We further explore how new software architectures might help
  application developers promote this style of end-user customization.
\end{abstract}

@@

These approaches have many differences, but they all share something in
common: an imperative programming model, with mutable variables,
conditionals and loops. End users ultimately must use these traditional
programming constructs to express their ideas, and the object of
interest is fundamentally a \emph{script}, a sequence of commands.

We have known for decades about an alternate approach: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the de facto standard in GUIs today, but when it comes
time to customize software, it is rarely to be found. In this work, we
ask: what would it look like to build a software customization interface
that deeply relies on direct manipulation? We take inspiration from
spreadsheets and visual database query interfaces
\citep{2020a, bakke2016}, which enable users to construct queries, edit
data, and compute derived values through direct manipulation. These
tools have successfully enabled millions of end users to compute with
data more easily than ever before.

In this paper we present a technique called \emph{table-driven
customization} which applies these ideas from visual query interfaces in
the context of software customization. We provide the end user with a
table view where they can directly see and manipulate the internal
structured data of an application. The table view and the original UI
serve as coordinated \emph{multiple representations} of the same
underlying state. Changes in the table view result in immediate
corresponding changes to the original user interface of the application,
enabling the user to customize an application with live feedback.

To demonstrate the viability of this approach, we have developed a
browser extension called Wildcard which implements table-driven
customization in the context of the Web. Wildcard works with existing
applications by using web scraping techniques to extract relevant state
into a table, and to reify changes back into the original application
UI. In Section~\ref{sec:examples} we present a many examples of using
Wildcard to implement useful customizations across real websites,
ranging from sorting lists of data to adding whole new features to
applications.

In Section~\ref{sec:architecture} we explain the architecture of the
Wildcard system, focusing on \emph{table adapters} which define the
semantics of how a data table is connected to some underlying state. We
explain how the abstractions of table-driven customization are general
enough to support many customizations beyond the ones we've built so far
in Wildcard. We also explore how applications themselves could be
re-architected to expose tabular data to end users, rather than relying
on web scraping techniques.

Table-driven customization embodies several design principles for
building customizable software systems, described in
Section~\ref{sec:design-principles}:

\begin{itemize}
\tightlist
\item
  \emph{Customization via direct manipulation}: End users should be able
  to customize an application by examining and modifying its data,
  rather than by issuing imperative commands to an API.
\item
  \emph{Third-party semantic wrappers}: Typically, ``black-box''
  customization tools, which don't rely on official extension APIs,
  resort to offering low-level APIs for customization. Instead, we
  suggest a third-party community of semantic wrappers, offering end
  users a more convenient semantic API for customization, even if it
  wasn't built in by the original developer.
\item
  \emph{Gentle slope}: Customization tools should offer a ``gentle
  slope'' from normal use to customization. We explore how our approach
  offers end users opportunities to start customizing software with
  almost no effort, which we hope can make customization a viable part
  of everyday software usage rather than a rare event.
\end{itemize}

Table-driven customization relates to existing work in many areas. Our
goals overlap with many software customization tools, in the browser and
on the desktop. Our methods overlap with direct manipulation interfaces
for working with data, including visual database query systems and
spreadsheets. We explore these connections and more in
Section~\ref{sec:related-work}.

@@

To concretely illustrate how table-driven customization works, here are
several examples of using the Wildcard browser extension, which is
implemented in Typescript and works with Chrome, Firefox, and Edge.

@@

@@

(\emph{Todo: drive this explanation via concrete Instacart example. Add
a figure.}) For example, one table adapter manages scraping data from
the DOM of a site, while another manages extracting data from AJAX
requests. Each table adapter produces a table of data, and the two are
joined together by a shared ID column to produce what looks like a
single table of data extracted from the site. Similarly, when the user
makes annotations on a site, these annotations are handled by a third
table adapter, which manages a data table stored in the browser's local
storage.

@@

\hypertarget{sec:design-principles}{%
\section{Design principles}\label{sec:design-principles}}

@@

\begin{itemize}
\tightlist
\item
  multiple representations
\item
  a discussion of
\end{itemize}

@@

Software customization tools typically fall into one of two categories:
black-box, or semantic.

@@

\emph{borrow from convivial paper's discussion of in-place toolchain}

@@

\begin{itemize}
\tightlist
\item
  our own workshop paper \citep{litt2020}. since then\ldots{}

  \begin{itemize}
  \tightlist
  \item
    fundamentally rearchitected around table adapters
  \item
    evaluated on many more websites
  \item
    more fully describing how the system works
  \end{itemize}
\item
  customization tools

@@

    \begin{itemize}
    \tightlist
    \item
      Mac Automator
    \item
      Zapier
    \item
      IFTTT
    \item
      These are weakly direct manipulation. But usually the thing you
      directly manipulate in these UIs is the \emph{script}: not the
      actual \emph{data} you want to operate on. VERY DIFFERENT.
    \end{itemize}
  \item
    desktop customization: Applescript, VBA, COM
  \end{itemize}
\item
  database GUIs:

  \begin{itemize}
  \tightlist
  \item
    Sieuferd \citep{bakke2016}
  \item
    Airtable \citep{2020a}
  \item
    other similar tools?
  \item
    Liu \& Jagadish: rules for a spreadsheet algebra for database
    queries \citep{liu2009}
  \item
    Shneiderman paper
  \item
    spreadsheets
  \end{itemize}
\item
  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
\item
  browser dev tools
\item
  Self, Smalltalk
\item
  multiple representations
\item
  Semantic Web
\end{itemize}

>>>>

  We describe Wildcard, a browser extension which implements
  table-driven customization in the context of web applications. Through
  concrete examples, we demonstrate that this paradigm can be used to
  create useful customizations for real applications. We share
  reflections from our usage of the Wildcard system, including its
  strengths and limitations relative to other customization approaches.
  We further explore how new software architectures might encourage
  application developers to promote this style of end-user
  customization.
\end{abstract}

@@

These approaches have many differences, but they all share something in
common: an imperative programming model, with mutable variables,
conditionals and loops. End users ultimately must use these traditional
programming constructs to express their ideas, and the object of
interest is a \emph{script}, a sequence of commands.

We have known for decades about an alternate approach: \emph{direct
manipulation} \citep{shneiderman1983}, where ``visibility of the object
of interest'' replaces ``complex command language syntax''. Direct
manipulation is the de facto standard in GUIs today, but when it comes
to customizing software, it is rarely to be found. In this work, we ask:
what would it look like to build a software customization interface that
deeply relies on direct manipulation? We take inspiration from
spreadsheets and visual database query interfaces
\citep{2020a, bakke2016}, which have successfully enabled end users to
construct queries, edit data, and compute derived values through direct
manipulation of structured data.

In this paper we present a technique called \emph{table-driven
customization} which applies these ideas from visual query interfaces in
the context of software customization. We augment an application's UI
with a table view, where the user can see and manipulate the structured
data inside an application. Changes in the table view result in
immediate corresponding changes to the original user interface of the
application, enabling the user to customize an application with live
feedback. The table view and the original UI serve as coordinated
\emph{multiple representations} of the same underlying state.

To demonstrate the viability of this approach, we have developed a
browser extension called Wildcard which implements table-driven
customization in the context of the Web. Wildcard is compatible with
existing unmodified applications because it uses web scraping techniques
to extract relevant state into a table and to reify changes in the
original application UI. In Section~\ref{sec:examples} we present a many
examples of using Wildcard to implement useful customizations across
real websites, ranging from sorting lists of data to adding whole new
features to applications.

In Section~\ref{sec:architecture} explain the \emph{table adapter}
abstraction, which allows many different types of underlying state to be
bidirectionally mapped to a table of data. We explore how table adapters
are general enough to support many customizations beyond the ones we've
built so far in Wildcard; we also discuss how applications could be
re-architected to expose tabular data directly to end users, rather than
relying on web scraping techniques.

In Section~\ref{sec:prototype} we discuss in more detail the
implementation of the Wildcard browser extension. We reflect on the
experience of using the tool to customize websites, and some of the
challenges from the perspective of programmers and end users.

Table-driven customization embodies several design principles for
building customizable software systems, discussed in
Section~\ref{sec:discussion}:

\begin{itemize}
\tightlist
\item
  \emph{Customization by direct manipulation}: End users should be able
  to customize an application by examining and modifying its data,
  rather than by writing imperative scripts.
\item
  \emph{Third-party semantic wrappers}: Typically, customization tools
  which don't rely on official extension APIs resort to offering
  low-level APIs for customization. Instead, we suggest a
  community-maintained library of semantic wrappers around existing
  applications, offering end users a more convenient semantic API for
  customization.
\item
  \emph{Gentle slope}: Customization tools should ideally offer a
  ``gentle slope'' from normal use to customization. We explore how our
  approach meets this goal by making some software customizations as
  easy as normal use of the software.
\end{itemize}

Table-driven customization relates to existing work in many areas. In
particular, our goals overlap with many software customization tools,
and our methods overlap with direct manipulation interfaces for working
with structured data, including visual database query systems and
spreadsheets. We explore these connections and more in
Section~\ref{sec:related-work}.

@@

To concretely illustrate the end user experience of table-driven
customization, here are several real examples of using the Wildcard
browser extension to customize websites.

@@

\begin{itemize}
\tightlist
\item
  basic idea: end user editing a table. but these edits have effects.
\item
  table adapter: component that presents an interface similar to a SQL
  table. but inside, it can do anything to produce that data, and can do
  anything with incoming edits.

  \begin{itemize}
  \tightlist
  \item
    out: table
  \item
    in: table
  \item
    in: other tables
  \end{itemize}
\end{itemize}

\begin{figure}
\hypertarget{fig:table-adapter}{%
\centering
\includegraphics[width=\columnwidth]{media/table-adapter.eps}
\caption{The table adapter architecture}\label{fig:table-adapter}
}
\end{figure}

@@

(\emph{Todo: drive this explanation via concrete example. Add a
figure.}) For example, one table adapter manages scraping data from the
DOM of a site, while another manages extracting data from AJAX requests.
Each table adapter produces a table of data, and the two are joined
together by a shared ID column to produce what looks like a single table
of data extracted from the site. Similarly, when the user makes
annotations on a site, these annotations are handled by a third table
adapter, which manages a data table stored in the browser's local
storage.

@@

\hypertarget{sec:evaluation}{%
\section{Evaluation}\label{sec:evaluation}}

\hypertarget{sec:discussion}{%
\section{Discussion}\label{sec:discussion}}

@@

\begin{itemize}
\tightlist
\item
  multiple representations \citep{ainsworth1999}
\item
  address Tchernavskij's point about contrasting with instrumental
  interaction
\end{itemize}

@@

Software customization tools typically fall into one of two categories:
third-party, or semantic.

@@

Originally came from \citep{maclean1990}

@@

This paper extends a workshop paper by Litt and Jackson \citep{litt2020}
which presented an earlier prototype of the Wildcard system. We build on
that work in this paper by rearchitecting the internal implementation of
the system around the table adapter abstraction, evaluating it on many
more websites, and by more fully characterizing the behavior and design
of the system.

\hypertarget{customization-tools}{%
\subsection{Customization tools}\label{customization-tools}}

\begin{itemize}
\tightlist
\item
  customization tools

@@

    \begin{itemize}
    \tightlist
    \item
      Mac Automator
    \item
      Zapier
    \item
      IFTTT
    \item
      These are weakly direct manipulation. But usually the thing you
      directly manipulate in these UIs is the \emph{script}: not the
      actual \emph{data} you want to operate on. VERY DIFFERENT.
    \end{itemize}
  \item
    desktop customization: Applescript, VBA, COM
  \end{itemize}
\end{itemize}

\hypertarget{visual-database-query-tools}{%
\subsection{Visual database query
tools}\label{visual-database-query-tools}}

\begin{itemize}
\tightlist
\item
  database GUIs:

  \begin{itemize}
  \tightlist
  \item
    Sieuferd \citep{bakke2016}
  \item
    Airtable \citep{2020a}
  \item
    other similar tools?
  \item
    Liu \& Jagadish: rules for a spreadsheet algebra for database
    queries \citep{liu2009}
  \item
    Shneiderman paper
  \item
    spreadsheets
  \end{itemize}
\end{itemize}

\hypertarget{other}{%
\subsection{Other}\label{other}}

\begin{itemize}
\tightlist
\item
  instrumental interaction, Scotty \citep{eagan2011}, Webstrates
\item
  browser dev tools
\item
  Self, Smalltalk
\item
  multiple representations
\item
  Semantic Web
\end{itemize}
